Holomorphic field theories and higher algebra

### 2024-09-14 draft.tex
reworking the intro to holo field theory section

A physical field theory is usually described in terms of a space of fields that forms some possibly nonlinear space, like a mapping space $\Map(M,X)$ or the space of connections ${\rm Conn}_G(M)$ on a manifold. 
There is a manifold, called spacetime in physics, where the fields are some kind of functions out of it;
in the examples just mentioned, it is denoted $M$.
In perturbative field theory, which describes the formal neighborhood of the theory at a point in such spaces, 
a field $\Phi$ is typically described by a smooth section of a vector bundle on spacetime.
The equations of motion that dictate the dynamics of the particular model are encoded by a series of partial differential equations of the form 
\[
\sum_{k \geq 1} D_k(\Phi,\ldots,\Phi) = 0 ,
\]
where $D_k$ is a $k$-ary polydifferential operator.
As an example, the scalar $\phi^4$-theory on a Riemannian $M$ is described by the equation
\[
\Delta \phi + \frac{1}{3!} \phi^3 = 0 
\]
where $\Phi = \phi$ is simply a smooth function on $M$.
These equations of motion are natural to the Riemannian geometry of the manifold $M$, 
and they can be derived from the Euler--Lagrange equations of the  action functional $\int_M \phi \Delta \phi \, \op{dvol}$.

As another example, consider perturbative Chern--Simons theory whose fields are a principal $G$-bundle $P$ on a smooth three-manifold $M$ equipped with a connection.
The equations of motion is the condition that the connection is flat 
\beqn
F_\nabla = 0 .
\eeqn
Perturbatively, near the trivial flat connection, the fields are simply $\lie{g} = \op{Lie}(G)$-valued one-forms $A \in \Omega^1(M, \lie{g})$.
The condition that perturbed connection $\d + A$ is flat is the Maurer--Cartan equation
\beqn
\d A + \frac12 [A,A] = 0.
\eeqn
These equations of motion are topological in the sense that they only depend on the smooth manifold~$M$.
When $\lie{g}$ is equipped with a non-degenerate, invariant, bilinear form $\kappa$ (for example, if $G$ is semi-simple there is a standard choice) then these equations of motion can be derived from the famous Chern--Simons action functional $\int_M CS(A)$ where
\beqn\label{eqn:CS}
CS(A) = \frac12 \kappa(A, \d A) + \frac16 \kappa(A, [A,A]) .
\eeqn

In a {\em holomorphic theory} one starts with a complex manifold $X$ as the spacetime.
%This means, in part, that the complexified tangent bundle of $X$ decomposes as a sum of vector bundles $\T_X \oplus \Bar{\T}_X$ where $\T_X$ (respectively $\Bar{T}_X$) is the holomorphic (respectively anti-holomorphic) tangent bundle.
%In local coordinates sections of $\T_X$ have the form $f^i \del / \del z_i$ whereas sections of $\Bar{\T}_X$ have the form $f^i \del / \del \zbar_i$. 
Globally, holomorphic theories are described like in the usual case by mapping spaces or spaces of connections.
The key feature in a holomorphic theory is that the equations of motion dictate that fields be holomorphic in an appropriate sense.
Perturbatively, where fields are sections of some (possibly graded) holomorphic vector bundle $V$, we can make this precise. 
Prescribing a holomorphic structure on a vector bundle $V$ is equivalent to the data of a $\dbar$-operator
\beqn
\dbar \colon \Omega^{0,q}(X, V) \to \Omega^{0,q+1}(X,V) 
\eeqn
where $\Omega^{0,q}(X,V)$ is the space of smooth sections of the vector bundle $\wedge^q\Bar{T}_X^* \otimes V$. 
Holomorphicity of $V$ boils down to the requirement that $\dbar \circ \dbar = 0$.
The Dolbeault complex of $V$ is
\beqn
\left(\Omega^{0,\bu}(X, V), \dbar \right) = \left(\oplus_{q \geq 0} \Omega^{0,q}(X, V)[-q], \dbar \right)
\eeqn
where (in the case that $V$ sits in degree zero) the space $\Omega^{0,q}(X,V)$ sits in degree $+q$.
A standard result in complex geometry is that the Dolbeault complex is a resolution for the sheaf of holomorphic sections of the holomorphic vector bundle $V$.
%In the case that $V$ is the trivial bundle $\ul{\CC}$ the Dolbeault complex $\Omega^{0,\bu}(X)$ is a resolution for the sheaf of holomorphic functions on $X$.

The fields of a holomorphic theory are given by elements of the Dolbeault complex $\Omega^{0,q}(X,V)$ for some $q$.
The equations of motion in a holomorphic theory take the form
\beqn
\dbar \Phi + \sum_{k \geq 1} D_k(\Phi,\ldots,\Phi) = 0 ,
\eeqn
where $D_k$ are $k$-ary {\em holomorphic} polydifferential operators.
This means that in local coordinates $D_k$ only depend on the holomorphic coordinates $\{z_i\}$ and involve only holomorphic derivatives $\del / \del z_i$.

For a simple example, consider the following holomorphic analog of Chern--Simons theory.
The fields are principal $G$-bundles on a three-dimensional \textit{complex} manifold $X$, and the equations of motion give the condition that the principal $G$-bundle is holomorphic.
Perturbatively, near the trivial $G$-bundle, the fields of this holomorphic theory are $\lie{g}$-valued $(0,1)$-forms $A \in \Omega^{0,1}(X, \lie{g})$.
The condition that $\d + A$ prescribes a holomorphic structure is again a Maurer--Cartan equation
\beqn
\dbar A + \frac12 [A,A] = 0 .
\eeqn
Like ordinary Chern--Simons theory, in order for this equation to arise from a principal of least action one needs to fix a non-degenerate, invariant, bilinear form $\kappa$ on $\lie{g}$.
Additionally, one needs to fix a holomorphic volume form $\Omega$ on $X$; in other words, $X$ is a Calabi--Yau manifold.
Then, the equations of motion is precisely the Euler--Lagrange equation for the action
\beqn
S_{hCS} (A) = \int_X \Omega \wedge CS(A) 
\eeqn
where $CS(A)$ is exactly as in \eqref{eqn:CS}.\footnote{Notice that due to the presence of the holomorphic volume form $\Omega$ only the $\dbar$ operator contributes to the kinetic term and so the action can be written as $\int_X \Omega \wedge \left(\frac12 \kappa( A ,\dbar A ) + \frac16 \kappa(A , [A,A]) \right)$.}

@@

@@

@@

\subsection{Perturbative field theories via derived deformations}

>>>>

In physics, a classical field theory is usually described in terms of a few pieces of data:
\begin{itemize}
\item a manifold $M$ that plays the role of ``spacetime,''
\item a space of fields, such as sections of some fiber bundle $F \to M$ or connections on a principal bundle on $M$, and
\item a system of PDE for the fields, known as the equations of motion.
\end{itemize}
These equations of motion are variational: 
they are the Euler-Lagrange equations of an action functional on the space a fields.
An action functional is written as an integral over $M$ of some polydifferential operator on the fields that takes values in densities of~$M$.
A few examples will illuminate this sketch.

\begin{eg}[Massless scalar field theory]
Let $M$ be a smooth manifold and equip it with a Riemannian metric~$g$.
Let $\Delta_g$ denote the associated Laplace-Beltrami operator, 
and let $\d{\rm vol}_g$ denote the associated volume form.
Let the space of fields be the smooth functions $C^\infty(M)$.
A smooth function $\phi$ is called, in physics, a {\em scalar} field.
The action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi \, \d{\rm vol}_g
\]
corresponds to the equation of motion $\Delta_g \phi = 0$. 
Thus this scalar field theory has the moduli space of {\em harmonic functions} as its space of solutions.
\end{eg}

The preceding example exhibits how a moduli space of central interest in differential geometry --- the harmonic functions --- appears naturally as a basic example in physics (the ``free massless scalar field'').
More complicated moduli spaces show up too.

\begin{eg}[$\phi^4$ theory]
As a small variant of the example just given,
consider the action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi + \frac{1}{4!} \phi^4\, \d{\rm vol}_g.
\]
Its equation of motion is the nonlinear PDE 
\[
\Delta_g \phi + \frac{1}{3!} \phi^3 = 0.
\] 
Thus this scalar field theory has a more complicated, nonlinear moduli space of solutions.
\end{eg}

We now describe an example of a theory whose space of solutions is a moduli space of interest in geometric topology.

\begin{eg}[Chern-Simons theory]
Pick a compact Lie group $G$ and equip it with an invariant metric,
which induces a $G$-invariant inner product $\kappa$ on its Lie algebra~$\fg$.
(The standard choice is the Killing form.)
Take an oriented smooth 3-manifold $M$, 
and consider the space of connections on the trivial principal $G$-bundle on $M$.
This space of fields is isomorphic to $\Omega^1(M, \fg)$, the $\fg$-valued one-forms, 
by viewing $\d + A$ as a connection where $A \in \Omega^1(M, \fg)$.
The equation of motion is the zero-curvature equation
\[
F_A = 0,
\]
which can be expressed equivalently as the Maurer-Cartan equation
\[
\d A + \frac{1}{2} [A,A] = 0.
\]
These equations arise as the Euler-Lagrange equation for the action functional
\beqn
\label{eqn:CS}
CS(A) = \frac12 \kappa(A, \d A) + \frac16 \kappa(A, [A,A]),
\eeqn
the integral of a Chern-Simons form.
Hence this theory is known as {\em Chern-Simons theory}.\footnote{Note that $M$ needs to be an oriented 3-manifold for this action functional to make sense.
The equation of motion exists for a principal bundle on any manifold of any dimension,
but it is only variational for oriented 3-manifolds.}
These equations do not depend on a metric on $M$ or any other geometric structure,
and hence define a ``topological'' field theory in the sense that they are independent of geometry.
This theory is interesting because the moduli space of solutions consists of the flat $G$-bundles on $M$,
a topic of deep interest in geometry and topology.
For physicists, it is interesting because it does not involve geometry, 
by contrast to earlier examples of field theories,
and yet exhibits rich phenomena (notably applications to the quantum Hall effect and 3-dimensional gravity).
\end{eg}

\begin{rmk}
In the Chern-Simons example, there is an important aspect we did not yet acknowledge:
we need to identify different solutions related by gauge transformations.
In other words, the correct moduli space is the quotient stack
\[
\left[ \{ A \,:\, F_A = 0\}/\Map(M,G) \right].
\]
The usual mode of description of a field theory, by specifying fields and an action functional (or equations of motion), 
must be supplemented by specifying whether to quotient by some group of symmetries of the space of solutions.
More generally, a field theory ought to present some kind of derived moduli stack.
\end{rmk}

We are now in a good position to offer an example of a {\em holomorphic field theory}:
the holomorphic analog of Chern-Simons theory.

%In a {\em holomorphic theory} one starts with a complex manifold $X$ as the spacetime.
%%This means, in part, that the complexified tangent bundle of $X$ decomposes as a sum of vector bundles $\T_X \oplus \Bar{\T}_X$ where $\T_X$ (respectively $\Bar{T}_X$) is the holomorphic (respectively anti-holomorphic) tangent bundle.
%%In local coordinates sections of $\T_X$ have the form $f^i \del / \del z_i$ whereas sections of $\Bar{\T}_X$ have the form $f^i \del / \del \zbar_i$. 
%Globally, holomorphic theories are described like in the usual case by mapping spaces or spaces of connections.
%The key feature in a holomorphic theory is that the equations of motion dictate that fields be holomorphic in an appropriate sense.
%Perturbatively, where fields are sections of some (possibly graded) holomorphic vector bundle $V$, we can make this precise. 
%Prescribing a holomorphic structure on a vector bundle $V$ is equivalent to the data of a $\dbar$-operator
%\beqn
%\dbar \colon \Omega^{0,q}(X, V) \to \Omega^{0,q+1}(X,V) 
%\eeqn
%where $\Omega^{0,q}(X,V)$ is the space of smooth sections of the vector bundle $\wedge^q\Bar{T}_X^* \otimes V$. 
%Holomorphicity of $V$ boils down to the requirement that $\dbar \circ \dbar = 0$.
%The Dolbeault complex of $V$ is
%\beqn
%\left(\Omega^{0,\bu}(X, V), \dbar \right) = \left(\oplus_{q \geq 0} \Omega^{0,q}(X, V)[-q], \dbar \right)
%\eeqn
%where (in the case that $V$ sits in degree zero) the space $\Omega^{0,q}(X,V)$ sits in degree $+q$.
%A standard result in complex geometry is that the Dolbeault complex is a resolution for the sheaf of holomorphic sections of the holomorphic vector bundle $V$.
%%In the case that $V$ is the trivial bundle $\ul{\CC}$ the Dolbeault complex $\Omega^{0,\bu}(X)$ is a resolution for the sheaf of holomorphic functions on $X$.
%
%The fields of a holomorphic theory are given by elements of the Dolbeault complex $\Omega^{0,q}(X,V)$ for some $q$.
%The equations of motion in a holomorphic theory take the form
%\beqn
%\dbar \Phi + \sum_{k \geq 1} D_k(\Phi,\ldots,\Phi) = 0 ,
%\eeqn
%where $D_k$ are $k$-ary {\em holomorphic} polydifferential operators.
%This means that in local coordinates $D_k$ only depend on the holomorphic coordinates $\{z_i\}$ and involve only holomorphic derivatives $\del / \del z_i$.

\begin{eg}[Holomorphic Chern-Simons theory]
\owen{This feels rather convoluted. Can we make it simpler?}
Pick a {\em complex} Lie group $G$, 
and equip its Lie algebra $\fg$ with a complex-linear, $G$-invariant nondegenerate symmetric bilinear form~$\kappa$.
Let $M$ be a three-dimensional \textit{complex} manifold
equipped with a holomorphic volume form $\Omega \in \Omega^{3,0}(M)$; 
in other words, $M$ is a Calabi--Yau manifold.
(This condition is analogous to being an oriented 3-manifold.)
Given a smooth principal $G$-bundle $P \to M$,
there is a moduli space of holomorphic structures on~$P$.
Suppose there exists at least one holomorphic structure $\cP \to M$,
which has an associated Dolbeault operator (or $\dbar$-connection) $\dbar_{\cP}$.
Any other holomorphic bundle structure determines a Dolbeault operator,
but our reference point $\dbar_{\cP}$ provides an isomorphism
\[
{\rm Conn}^\dbar_G(P) \cong \Omega^{0,1}(M, {\rm ad}(P)),
\]
i.e., we express another $\dbar$-connection as $\dbar_{\cP} + A$ where $A$ is a $(0,1)$-form valued in the adjoint bundle.
The equation of motion is again a Maurer--Cartan equation
\beqn
\dbar_{\cP} A + \frac12 [A,A] = 0 ,
\eeqn
or equivalently the $(0,2)$-term of the curvature vanishes.
This PDE is the Euler--Lagrange equation for the action
\beqn
S_{hCS} (A) = \int_X \Omega \wedge CS(A) 
\eeqn
where $CS(A)$ is exactly as in~\eqref{eqn:CS}.\footnote{Notice that due to the presence of the holomorphic volume form $\Omega$, only the $\dbar A$ component of $\d A$ contributes to the integrand of the action functional, because $\Omega \wedge \partial A$ vanishes.} 
As in the case of Chern-Simons theory, one must quotient by the appropriate group of gauge transformations to obtain the correct moduli space, 
which is here the moduli stack of holomorphic principal $G$-bundles whose underlying smooth structure is given by $P \to M$.
\end{eg}

For a systematic introduction to classical field theory in a style amenable to a complex geometer,
we suggest \cite{FreedDeligne},
although there are many other sources, such as \cite{}.
For a detailed development of field theory in dialogue with derived geometry,
see \cite{Steffens},
and for a motivational discussion, likely more accessible to physicists, see \cite{AlfonsoYoung?}.
In Section~\ref{sec: pert dfn} we offer a sharp definition well-suited to perturbative field theory 
(i.e., studying formal deformations of a fixed solution to the equations of motion),
which we will use to produce examples of higher algebras.

@@

In this section we describe several examples of holomorphic field theories,
close to the style we used for holomorphic Chern-Simons theory.

@@

\owen{Somewhere cite Donaldson and Thomas.}

@@

\subsection{Perturbative field theories via derived deformations}
\label{sec: pert dfn}
